---
title: "LAB_02_Map_Making_Assignment"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

#Study Area Description

This is a map showing part of Auburn University (study area), located in Auburn, Lee County, Alabama. Auburn University is a land-grant institution in the Southeastern US located at the fault line of two distinct ecological regions: the Southeastern Plains and the Piedmont. The northern portion of the city of Auburn is best classified as part of the Southern Outer Piedmont, a subset of the Piedmont ecological region. The Southern Outer Piedmont region has lower elevation and lower levels of annual precipitation. Although the gently sloping hills of the region are described as mainly forested throughout history, the study area is currently a predominantly urban area. Auburn University’s campus is well established, with many large buildings, athletic complexes, roads, and other impervious surfaces.
With increasing human population and urbanization, other land cover classes such as forest and agricultural lands have to be converted to make space for urban expansions. These land cover changes have significant impact on biodiversity, including birds.This study was conducted in part of Auburn University to survey birds in different land cover classes and understand how bird population and distribution has changed in these areas as a function of land cover changes.
The study area map shows points that were sampled for the years 2000 and 2024 and the elevation of the study area. NAIP imagery (1m resolution) was downloaded from the United States Geological Survey's (USGS) Earth Explorer. This was plotted on the study area to show the various land cover classes in the study area and their distribution. The Naip imagery was used to select and create the sampling points in ArcGIS Pro to span a latitudinal gradient across Auburn and to cover major land cover classes (Forest, Grassland, Water, Urban areas) in the study area. 

#Datasets used in this assignment

##Vector Elements

1. Lee_County, Alabama boundary Shapefile; downloaded from DATA.GOV website (https://catalog.data.gov/dataset/tiger-line-shapefile-2019-nation-u-s-current-county-and-equivalent-national-shapefile) and clipped to study area using ArcGIS Pro.

2. Lee_County_Points shapefile; This dataset was created in ArcGIS Pro. Random points were created as sampling points within the study area for two chosen years (2000 and 2024).

##Raster Elements

3. Naip; This is a Naip imagery that was downloaded for the study area. The dataset has a spatial resolution of 1m and four spectral bands (RGB and NIR).

4. Elevation dataset

#Loading required packages

```{r}
require(tidyverse)
require(sf)
require(tigris)
require(geodata)
require(terra)
require(colorspace)
require(tidyterra)
require(ggnewscale)
require(cowplot)
```

#Zoomed Out MaP
*This is going to be a map showing the study area and sampling points within Lee county, Alabama.* 

Step 1. Adding the Lee County boundary shapefile. 

```{r}
Lee_County = vect("/vsicurl/https://github.com/sza0209/Spatial-Ecology/raw/main/Lab_02/Lee%20county.shp")
plot(Lee_County)
```

Step 2. Adding the Lee County Points shapefile. 

```{r}
Lee_County_Points = st_read("/vsicurl/https://github.com/sza0209/Spatial-Ecology/raw/main/Lab_02/Points2.shp")
plot(st_geometry(Lee_County_Points))
```

Step 3. Plotting both the Lee_County and Lee_County_Points together. To do this, first, I will transform the Lee_County_Points shapefile to the same coordinate system as the Lee_County shapefile to ensure the points fall in the same place.

```{r}
# Plotting all shapefiles together to form the zoomed out map of the study area
Lee_County_Points = st_transform(Lee_County_Points, st_crs(Lee_County))
plot(Lee_County, col='yellow')
plot(st_geometry(Lee_County_Points), col='black', add=T)
```

Step 4. Creating a bounding box to show the extent of area encompassed by the sampling points to be used as the study area

```{r}
#Extracting x and y coordinates of the bounding box using the points
studyArea = st_bbox(Lee_County_Points)
studyArea

#Convertig it to a spatial sfc object
studyArea = st_as_sfc(studyArea)

plot(studyArea)
```

Step 5. Plotting the *Zoomed Out* map showing Lee_County, the study area and the sampling points.

```{r}
plot(Lee_County)
plot(studyArea, col='yellow', add=T)
plot(st_geometry(Lee_County_Points), add=T)
```

Step 6. Plotting the Zoomed Out Map.

```{r}
##Zoomed out map
zoomedOut = ggplot()+
  geom_sf(data = Lee_County)+
  geom_sf(data = studyArea, fill=NA, color='red')+
  geom_sf(data = Lee_County_Points)

zoomedOut
```

Step 7. Removing gridlines, axis text, axis tick marks and frame around the graph to produce the final *Zoomed Map*.

```{r}
zoomedOut = zoomedOut +
  theme_bw()+ 
  theme(panel.grid=element_blank(), 
        axis.text=element_blank(), 
        axis.ticks=element_blank(), 
        panel.border=element_blank()) 

zoomedOut
```


#Zoomed In MaP
*This is going to be a map showing a NAIP imagery and elevation of study area and sampling points*

Step 1. Downloading elevation data for the United States and cropping it to the state of Alabama.

```{r}
Alabama = states() %>% 
  filter(NAME=='Alabama')

plot(st_geometry(Alabama))
```

```{r}
usaElev = elevation_30s(country='USA', path=tempdir())


Alabama_Elev = st_transform(Alabama, st_crs(usaElev))
AlabamaElev = crop(usaElev, Alabama)

plot(AlabamaElev)
plot(st_geometry(Alabama), add=T)
```
```{r}
AlabamaElev = mask(AlabamaElev, Alabama)
plot(AlabamaElev)
```

Step 2. Zooming into the study area.

```{r}
studyArea = st_transform(studyArea, st_crs(Alabama))
Elevation = crop(AlabamaElev, studyArea)
plot(Elevation, col=sequential_hcl(13, "Terrain"))
```

Step 3; Importing the Naip imagery of study area. This shows the study area in natural colors (RGB) just as the naked eye sees it at a spatial resolution of 1m.

```{r}
Naip = rast('C:/Users/jjv0016/OneDrive - Auburn University/Teaching/Spatial Ecology/Lab2/submissions/abubakarsinkakhadijah_4040919_236807945_NAIP Imagery/AuburnNAIP2014/m_3208529_ne_16_1_20151014.tif')

studyArea = st_transform(studyArea, st_crs(Naip))
Naip = crop(Naip, studyArea)
Naip[is.na(Naip)] = 0
plot(Naip)
plotRGB(Naip)
```

Step 4. Checking and converting the projections of the datasets.

```{r}
crs(Elevation, proj=T)
crs(Lee_County, proj=T)
crs(Lee_County_Points, proj=T)
crs(Naip, proj=T)
```

```{r}
Lee_County = project(Lee_County, crs(Naip, proj=T))
Elevation = project(Elevation, crs(Naip, proj=T))
```

```{r}
crs(Elevation, proj=T)
crs(Lee_County, proj=T)
crs(Lee_County_Points, proj=T)
crs(Naip, proj=T)
```

Step 5. Plotting the elevation of the study area in black and white gradient.

```{r}
zoomedIn = ggplot()+
  geom_spatraster(data = Elevation)+
  scale_fill_gradient(low='white', high='black', labs(fill='Elevation'))

zoomedIn
```

Step 6. Overlaying the Naip imagery, elevation dataset and sampling points to plot the final zoomed in map.

```{r}
tmp = Lee_County_Points %>%
  mutate(SamplingYr = as.character(SamplingYr))

zoomedIn = zoomedIn + 
  new_scale_fill()+
  geom_spatraster_rgb(data=Naip, alpha=0.5)+
  # scale_color_manual(values=c('white', 'red'))+
  geom_sf(data=tmp, aes(color=SamplingYr))

zoomedIn
```

Step 7. Plotting the final zoomed in map.

```{r}
zoomedIn = zoomedIn+
  new_scale_fill()+
  theme_bw()+
  theme(panel.grid=element_blank(),
        axis.text.x=element_text(angle=90, vjust=0.5, size=6))

zoomedIn
```

Step 8; Adding the scale bar and north arrow.

```{r}
map <- zoomedIn+
  ggspatial::annotation_scale(
    location = "tr",
    bar_cols = c("grey60", "white"),
    text_family = "ArcherPro Book"
  ) +
  ggspatial::annotation_north_arrow(
    location = "tr", which_north = "true",
    pad_x = unit(0.5, "in"), pad_y = unit(0.5, "in"),
    style = ggspatial::north_arrow_nautical(
      fill = c("grey40", "white"),
      line_col = "grey20",
      text_family = "ArcherPro Book"
    )
  )

map
```

Final step; Piecing together the zoomed in and zoomed out maps to create the final study area map.

```{r}

plot_grid(zoomedOut, map, rel_widths=c(0.5, 1))

```

# Grade

$\color{red}{\text{Nice work on this, Sinka.}}$