---
title: "Lab 4 Assignment - Landcover"
output:
  html_document:
    df_print: paged
---



```{r}
# Load your required R packages here.
rm(list=ls())
require(Voss)
require(terra)
require(tidyverse)
require(FedData)
require(sf)
require(tidyterra)
require(landscapemetrics)
# require(ggplot2)
```

## Challenge 1 (4 points)

**The landscapemetrics package has functions for calculating 12 patch-level metrics. Calculate all 12 of these for every forest patch in our nlcdSimple raster (using the 8-direction rule). This will result in 12 unique values associated with each forest patch. Use the chart.Correlation() function from the PerformanceAnalytics package to examine the correlation among all 12 of these variables. What patterns do you notice, and what do they tell you about the uniqueness or redundancy of the patch-level metrics you calculated?**

```{r}
studyArea = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week4/studyArea.shp')
nlcd = get_nlcd(studyArea, label='AlLandscape', year=2011)


plot(nlcd)

nlcdSimple = nlcd
nlcdSimple[nlcdSimple==11] = 1 
nlcdSimple[nlcdSimple %in% c(21, 22, 23, 24)] = 2
nlcdSimple[nlcdSimple %in% c(31, 52)] = 3 
nlcdSimple[nlcdSimple %in% c(41,42,43)] = 4 
nlcdSimple[nlcdSimple == 71] = 5 
nlcdSimple[nlcdSimple %in% c(81,82)] = 6 

# Naming the categories
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcdSimple = categories(nlcdSimple, value=tmp)

#Plotting the new raster
ggplot(nlcdSimple, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))

forest = nlcdSimple %>% 
  setValues(NA)
  
forest[nlcdSimple ==4] = 1

plot(forest)

#Calculating the 12 metrics
forestPatchId = patches(forest, directions=8, zeroAsNA=T, allowGaps=F)

plot(forestPatchId)

## Area 
patchArea = lsm_p_area(forest, directions=8)

patchArea

## Core
patchCore = lsm_p_core(forest, directions=8)

patchCore

## Enn
patchEnn = lsm_p_enn(forest, directions=8)

patchEnn

## Perimeter
patchPerim = lsm_p_perim(forest, directions=8)

patchPerim

## Para
patchPara = lsm_p_para(forest, directions=8)

patchPara

## Cai
patchCai = lsm_p_cai(forest, directions=8)

patchCai

## Circle
patchCircle = lsm_p_circle(forest, directions=8)

patchCircle

## Contig
patchContig = lsm_p_contig(forest, directions=8)

patchContig

## Frac
patchFrac = lsm_p_frac(forest, directions=8)

patchFrac

## Gyrate
patchGyrate = lsm_p_gyrate(forest, directions=8)

patchGyrate

## Ncore
patchNcore = lsm_p_ncore(forest, directions=8)

patchNcore

## Shape
patchShape = lsm_p_shape(forest, directions=8)

patchShape


# Combining metric values into a single data frame
Patch_metrics <- data.frame(
  Area = patchArea$value,
  CoreAreaIndex = patchCai$value,
  Circle = patchCircle$value,
  ContiguityIndex = patchContig$value,
  Core = patchCore$value,
  EuclideanNNDistance = patchEnn$value,
  FractalDimensionIndex = patchFrac$value,
  RadiusOfGyration = patchGyrate$value,
  NumberOfCoreAreas = patchNcore$value,
  PerimeterAreaRatio = patchPara$value,
  Perimeter = patchPerim$value,
  Shape = patchShape$value
)
 
# Performing correlation
PerformanceAnalytics::chart.Correlation(Patch_metrics, histogram=TRUE, pch=19)
```

Answer to Challenge 1.
There are very high correlations between the area and core (1.00), area and perimeter (0.95), core and perimeter (0.92) and shape and dimension (0.95) metrics indicating redundancy. On the other hand, low correlation values are observed between Contiguity Index and Perimeter Area Ratio (-0.99) suggesting uniqueness.

$\color{red}{\text{The latter part of what you said is not exactly true. Two metrics are uncorrelated when they have a Pearson's correlation coefficient of 0. A value of -0.99 actually indicates nearly complete negative correlation. +3.5}}$



## Challenge 2 (4 points)

**In our lab, we used the 8-direction or "queen" rule to delineate patches. Using the nlcdSimple raster we created, explore the differences in patch characteristics if you were to use the 4-direction or "rook" rule for delineating patches. Calculate the following class-level metrics for forest cover for both the queen and rook patch delineation rules: number of patches, mean patch size, standard deviation in patch size, nearest-neighbor distance, and total edge length. What patterns do you notice? When might it be appropriate to use one rule vs. the other?**

```{r}
# Using the 8-cell neighbor rule

# Number of Patches
NumPatch8 = lsm_c_np(forest, directions=8)
NumPatch8
 
# Mean Patch Size
MeanPatch8 = lsm_c_area_mn(forest, directions=8)
MeanPatch8
 
# Standard Deviation in Patch Size
SDPatch8 = lsm_c_area_sd(forest, directions=8)
SDPatch8
 
# Mean Nearest-neigbor Distance
NNDistance8 = lsm_c_enn_mn(forest, directions=8)
NNDistance8

# Total edge length
TEL8 = lsm_c_te(forest, directions=8)
TEL8

# Using the 4-cell neighbor rule

# Number of Patches
NumPatch4 = lsm_c_np(forest, directions=4)
NumPatch4
 
# Mean Patch Size
MeanPatch4 = lsm_c_area_mn(forest, directions=4)
MeanPatch4
 
# Standard Deviation in Patch Size
SDPatch4 = lsm_c_area_sd(forest, directions=4)
SDPatch4
 
# Mean Nearest-neigbor Distance
NNDistance4 = lsm_c_enn_mn(forest, directions=4)
NNDistance4

# Total edge length
TEL4 = lsm_c_te(forest, directions=4)
TEL4
```

Answer to Challenge 2.
The rook rule recorded the highest number of patches (67), whereas the queen rule recorded 49. The queen rule might often result in less and larger patches compared to the rook rule because it considers diagonals. The queen rule recorded the higher values in mean patch size, standard deviation in patch size and nearest-neighbor distance. Queen rule tends to result in larger mean patch sizes and potentially more variability. Nearest-Neighbor Distance may differ between the queen and rook rules, reflecting the spatial arrangement of patches. Both rules however recorded total edge values of 0.

When to use one rule over the other depends on the spatial characteristics of your data and the specific ecological or landscape context. The queen rule is often more appropriate when considering landscape connectivity and habitat fragmentation, while the rook rule may be suitable for a simpler representation of spatial relationships. 

$\color{red}{\text{Nice. Ultimately it boils down to "what's your question?". +4}}$


## Challenge 3 (4 points)


**Using the same zoomed-in study area that we used in the lab, download NLCD raster data for the years 2001 and 2019 and simplify these rasters into 6 landcover categories (wet, developed, barren, forest, grassland, and agriculture). Plot these two rasters. What are some of the initial changes you notice between 2001 and 2019?**

```{r}
#Getting NLCD data for the year 2001
nlcd_2001 = get_nlcd(studyArea, label='AlLandscape', year=2001)

plot(nlcd_2001)

nlcd_2001[nlcd_2001==11] = 1 
nlcd_2001[nlcd_2001 %in% c(21, 22, 23, 24)] = 2 
nlcd_2001[nlcd_2001 %in% c(31, 52)] = 3 
nlcd_2001[nlcd_2001 %in% c(41,42,43)] = 4 
nlcd_2001[nlcd_2001 == 71] = 5 
nlcd_2001[nlcd_2001 %in% c(81,82)] = 6 


tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcd_2001 = categories(nlcd_2001, value=tmp)


ggplot(nlcd_2001, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))


#Getting NLCD data for the year 2019
nlcd_2019 = get_nlcd(studyArea, label='AlLandscape', year=2019)

plot(nlcd_2019)

nlcd_2019[nlcd_2019==11] = 1 
nlcd_2019[nlcd_2019 %in% c(21, 22, 23, 24)] = 2 
nlcd_2019[nlcd_2019 %in% c(31, 52)] = 3 
nlcd_2019[nlcd_2019 %in% c(41,42,43)] = 4 
nlcd_2019[nlcd_2019 == 71] = 5 
nlcd_2019[nlcd_2019 %in% c(81,82)] = 6 


tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcd_2019 = categories(nlcd_2019, value=tmp)


ggplot(nlcd_2019, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))
```

Answer to Challenge 3.
Major land use changes can be observed from the rasters indicating shifts in land use patterns. Forest patches can be seen to reduce drastically within the study area from 2001 to 2019. Most of these patches have been lost to open areas, grassland areas and agricultural lands. Developed areas have also increased in the site indicating urbanization/ developments.


**Quantify this at the class level by calculating and reporting the changes in (1) the total amount of each land cover type (2) mean patch size for each land cover type, and (3) mean nearest neighbor distance for each cover type between the years 2011 and 2019. Give a short description of how you interpret the changes in these values.**


```{r}
classCats = data.frame('class' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))

# Class metrics for the 2001 nlcd

#Total area of each land cover type
lsm_c_ca(nlcd_2001, directions=8) %>% 
  left_join(classCats, by='class')

# Mean patch size for each land cover type
lsm_c_area_mn(nlcd_2001, directions=8) %>% 
  left_join(classCats, by='class')

#Total area of each land cover type
lsm_c_enn_mn(nlcd_2001, directions=8) %>% 
  left_join(classCats, by='class')


# Class metrics for the 2019 nlcd

#Total area of each land cover type
lsm_c_ca(nlcd_2019, directions=8) %>% 
  left_join(classCats, by='class')

# Mean patch size for each land cover type
lsm_c_area_mn(nlcd_2019, directions=8) %>% 
  left_join(classCats, by='class')

#Total area of each land cover type
lsm_c_enn_mn(nlcd_2019, directions=8) %>% 
  left_join(classCats, by='class')
```
Answer

Total Amount of Each Land Cover Type: Developed areas increased from 56.70 in 2001 to 58.05 2019. Forest area decreased significantly from 997.74 to 584.28, potentially indicating deforestation or changes in land use land cover. Agriculture areas also increased from 500.40 to 627.48.Wetland areas remained relatively the same in both years. Open areas increased significantly from 22.32 to 181.71.

Mean Patch Size: Mean patch size for developed areas decreased slightly. Forest mean patch size decreased, suggesting smaller, less contiguous forested areas. Agriculture mean patch size increased, indicating larger agricultural patches.

Mean Nearest Neighbor Distance: Generally, mean nearest neighbor distance decreased acrossed all classes, suggesting increased proximity between patches.



**Quantify these changes at the landscape level by calculating and reporting on changes in the (1) Shannon diversity and (2) Shannon evenness of the landscapes at the different time points. Give a short description of how you interpret the changes in these values.**

```{r}
# Shannon diversity for both nlcd 2001 and 2019 maps
lsm_l_shdi(nlcd_2001)
lsm_l_shdi(nlcd_2019)

# Shannon evenness for both nlcd 2001 and 2019 maps
lsm_l_shei(nlcd_2001)
lsm_l_shei(nlcd_2019)
```

Answer

Shannon Diversity (shdi); Shannon Diversity increased from 2001 (~0.95) to 2019 (~1.34). Shannon diversity measures the variety and abundance of different land cover classes, with higher diversity values indicating a more diverse landscape in terms of land cover types. With an increment in shdi from 2001 to 2019, this suggests an increase in the variety and abundance of different land cover types, potentially indicating landscape complexity or ecological changes.

Shannon Evenness (shei); Shannon Evenness increased from 2001 (~0.52) to 2019 (~0.74). Shannon evenness measures the equitability or uniformity in the distribution of land cover classes, with higher evenness values indicating a more even distribution of land cover types.
An increment from 2001 to 2019 suggests a more balanced distribution of land cover classes in the landscape, indicating a more evenly distributed or balanced landscape.

$\color{red}{\text{Perfect. +4}}$


## Challenge 4 (4 points)

**Use the voss2d() function to simulate a surface where g = 7 and H = 0.5. From that, create 9 'landscapes' with 10%, 20%, 30%, ..., 90% threshold values. The '1' values here can represent anything your imagination comes up with. It could be forest cover, cover of some other land cover type, bodies of water, temperatures above a threshold, etc. I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes and comment on what patterns you see changing as the value increases from a 10% cover to 90% cover.**

```{r}
set.seed(23)

vossModel = voss2d(g=7, H=0.5)
vossModel = rast(vossModel$z)

plot(vossModel)

# 10% cover
threshold10 = quantile(as.matrix(vossModel), prob=0.1)
voss10 = ifel(vossModel > threshold10, 0, 1)
plot(voss10)

# 20% cover
threshold20 = quantile(as.matrix(vossModel), prob=0.2)
voss20 = ifel(vossModel > threshold20, 0, 1)
plot(voss20)

# 30% cover
threshold30 = quantile(as.matrix(vossModel), prob=0.3)
voss30 = ifel(vossModel > threshold30, 0, 1)
plot(voss30)

# 40% cover
threshold40 = quantile(as.matrix(vossModel), prob=0.4)
voss40 = ifel(vossModel > threshold40, 0, 1)
plot(voss40)

# 50% cover
threshold50 = quantile(as.matrix(vossModel), prob=0.5)
voss50 = ifel(vossModel > threshold50, 0, 1)
plot(voss50)

# 60% cover
threshold60 = quantile(as.matrix(vossModel), prob=0.6)
voss60 = ifel(vossModel > threshold60, 0, 1)
plot(voss60)

# 70% cover
threshold70 = quantile(as.matrix(vossModel), prob=0.7)
voss70 = ifel(vossModel > threshold70, 0, 1)
plot(voss70)

# 80% cover
threshold80 = quantile(as.matrix(vossModel), prob=0.8)
voss80 = ifel(vossModel > threshold80, 0, 1)
plot(voss80)

# 90% cover
threshold90 = quantile(as.matrix(vossModel), prob=0.9)
voss90 = ifel(vossModel > threshold90, 0, 1)
plot(voss90)
```

Answer
With a threshold of 10%, you notice a few scattered patches of forest cover (assuming our value of 1 represents forest).More patches are connected, forming larger forested areas with 20% threshold. There is further consolidation of forested areas with increased connectivity as the threshold increases.With a threshold of 70%, you notice a few isolated non-forest patches within a predominantly forested landscape. By 90% almost the entire landscape is forested, with only tiny isolated non-forest areas.

**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is threshold value and y-axis is calculated metric). Briefly describe why you chose these 3 metrics and how they change with increasing cover.**

```{r}
# Metric; Class area, Mean of core area, edge density

#Class area for the nine landscapes
CA_10 = lsm_c_ca(voss10, directions=8) %>% 
  filter(class==1)
CA_10 <- CA_10$value

CA_20 = lsm_c_ca(voss20, directions=8) %>% 
  filter(class==1)
CA_20 <- CA_20$value

CA_30 = lsm_c_ca(voss30, directions=8) %>% 
  filter(class==1)
CA_30 <- CA_30$value

CA_40 = lsm_c_ca(voss40, directions=8) %>% 
  filter(class==1)
CA_40 <- CA_40$value

CA_50 = lsm_c_ca(voss50, directions=8) %>% 
  filter(class==1)
CA_50 <- CA_50$value

CA_60 = lsm_c_ca(voss60, directions=8) %>% 
  filter(class==1)
CA_60 <- CA_60$value

CA_70 = lsm_c_ca(voss70, directions=8) %>% 
  filter(class==1)
CA_70 <- CA_70$value

CA_80 = lsm_c_ca(voss80, directions=8) %>% 
  filter(class==1)
CA_80 <- CA_80$value

CA_90 = lsm_c_ca(voss90, directions=8) %>% 
  filter(class==1)
CA_90 <- CA_90$value

#Plotting
Class_Area <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  CA_Values = c(CA_10, CA_20, CA_30,
                 CA_40, CA_50, CA_60,
                 CA_70, CA_80, CA_90)
)
 
ggplot(Class_Area, aes(x = Threshold, y = CA_Values)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Class Area vs Threshold Value",
       x = "Threshold",
       y = "Class Area")
  
# Edge density
ED_10 = lsm_c_ed(voss10, directions=8) %>% 
  filter(class==1)
ED_10 <- ED_10$value

ED_20 = lsm_c_ed(voss20, directions=8) %>% 
  filter(class==1)
ED_20 <- ED_20$value

ED_30 = lsm_c_ed(voss30, directions=8) %>% 
  filter(class==1)
ED_30 <- ED_30$value

ED_40 = lsm_c_ed(voss40, directions=8) %>% 
  filter(class==1)
ED_40 <- ED_40$value

ED_50 = lsm_c_ed(voss50, directions=8) %>% 
  filter(class==1)
ED_50 <- ED_50$value

ED_60 = lsm_c_ed(voss60, directions=8) %>% 
  filter(class==1)
ED_60 <- ED_60$value

ED_70 = lsm_c_ed(voss70, directions=8) %>% 
  filter(class==1)
ED_70 <- ED_70$value

ED_80 = lsm_c_ed(voss80, directions=8) %>% 
  filter(class==1)
ED_80 <- ED_80$value

ED_90 = lsm_c_ed(voss90, directions=8) %>% 
  filter(class==1)
ED_90 <- ED_90$value

#Plotting
Edge_Density <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  ED_Values = c(ED_10, ED_20, ED_30,
                 ED_40, ED_50, ED_60,
                 ED_70, ED_80, ED_90)
)
 
ggplot(Edge_Density, aes(x = Threshold, y = ED_Values)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Edge density vs Threshold Value",
       x = "Threshold",
       y = "Edge density")

# Mean of core area
MCA_10 = lsm_c_core_mn(voss10, directions=8) %>% 
  filter(class==1)
MCA_10 <- MCA_10$value

MCA_20 = lsm_c_core_mn(voss20, directions=8) %>% 
  filter(class==1)
MCA_20 <- MCA_20$value

MCA_30 = lsm_c_core_mn(voss30, directions=8) %>% 
  filter(class==1)
MCA_30 <- MCA_30$value

MCA_40 = lsm_c_core_mn(voss40, directions=8) %>% 
  filter(class==1)
MCA_40 <- MCA_40$value

MCA_50 = lsm_c_core_mn(voss50, directions=8) %>% 
  filter(class==1)
MCA_50 <- MCA_50$value

MCA_60 = lsm_c_core_mn(voss60, directions=8) %>% 
  filter(class==1)
MCA_60 <- MCA_60$value

MCA_70 = lsm_c_core_mn(voss70, directions=8) %>% 
  filter(class==1)
MCA_70 <- MCA_70$value

MCA_80 = lsm_c_core_mn(voss80, directions=8) %>% 
  filter(class==1)
MCA_80 <- MCA_80$value

MCA_90 = lsm_c_core_mn(voss90, directions=8) %>% 
  filter(class==1)
MCA_90 <- MCA_90$value

#Plotting
Mean_Core_Area <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  MCA_Values = c(MCA_10, MCA_20, MCA_30,
                 MCA_40, MCA_50, MCA_60,
                 MCA_70, MCA_80, MCA_90)
)
 
ggplot(Mean_Core_Area, aes(x = Threshold, y = MCA_Values)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Mean Core Area vs Threshold Value",
       x = "Threshold",
       y = "Mean Core Area")
```

Answer

Class-level metrics used to capture changes; Class Area, Edge Density and Mean Core Area.

Why These Metrics:

Class Area: Provides information on the total area covered by the forest class in the landscape. It helps quantify the proportion of the landscape covered by the simulated forest.

Edge Density: Represents the amount of edge between forest and non-forest areas. Higher edge density can indicate more fragmented landscapes.

Mean Core Area: Reflects the average size of the core areas (interior patches of the forest).It helps understand the interior habitat quality, which can be important for species that require large, undisturbed areas.
 
How They Change:

Class Area: there is an increasing trend in the forest class (value 1) as the threshold increases, indicating a larger proportion of the landscape covered by forest.

Edge Density: there was an initial increase with lower threshold values as forest patches become more fragmented. This plateaued and started to decrease at higher thresholds as large connected forest areas dominate.

Mean Core Area: there is an increase in mean core area values as forest patches consolidate. 

$\color{red}{\text{Awesome. +4}}$


## Challenge 5 (4 points)

**Use the voss2d() function to simulate 9 surfaces. In each one, g should be 7, but vary the value of H from 0.1 to 0.9. Create a unique landscape from each of these with a threshold value of 30% cover. Again, the 1 values can represent anything you desire, and again I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes and comment on what patterns you see changing as the H value increases from 0.1 to 0.9.**

```{r}
set.seed(23)

# Lanscape with H value of 0.1
vossModel_0.1 = voss2d(g=7, H=0.1)
vossModel_0.1 = rast(vossModel_0.1$z)
threshold30 = quantile(as.matrix(vossModel_0.1), prob=0.3)
vossModel_0.1 = ifel(vossModel_0.1 > threshold30, 0, 1)
plot(vossModel_0.1)

# Lanscape with H value of 0.2
vossModel_0.2 = voss2d(g=7, H=0.2)
vossModel_0.2 = rast(vossModel_0.2$z)
threshold30 = quantile(as.matrix(vossModel_0.2), prob=0.3)
vossModel_0.2 = ifel(vossModel_0.2 > threshold30, 0, 1)
plot(vossModel_0.2)

# Lanscape with H value of 0.3
vossModel_0.3 = voss2d(g=7, H=0.3)
vossModel_0.3 = rast(vossModel_0.3$z)
threshold30 = quantile(as.matrix(vossModel_0.3), prob=0.3)
vossModel_0.3 = ifel(vossModel_0.3 > threshold30, 0, 1)
plot(vossModel_0.3)

# Lanscape with H value of 0.4
vossModel_0.4 = voss2d(g=7, H=0.4)
vossModel_0.4 = rast(vossModel_0.4$z)
threshold30 = quantile(as.matrix(vossModel_0.4), prob=0.3)
vossModel_0.4 = ifel(vossModel_0.4 > threshold30, 0, 1)
plot(vossModel_0.4)

# Lanscape with H value of 0.5
vossModel_0.5 = voss2d(g=7, H=0.5)
vossModel_0.5 = rast(vossModel_0.5$z)
threshold30 = quantile(as.matrix(vossModel_0.5), prob=0.3)
vossModel_0.5 = ifel(vossModel_0.5 > threshold30, 0, 1)
plot(vossModel_0.5)

# Lanscape with H value of 0.6
vossModel_0.6 = voss2d(g=7, H=0.6)
vossModel_0.6 = rast(vossModel_0.6$z)
threshold30 = quantile(as.matrix(vossModel_0.6), prob=0.3)
vossModel_0.6 = ifel(vossModel_0.6 > threshold30, 0, 1)
plot(vossModel_0.6)

# Lanscape with H value of 0.7
vossModel_0.7 = voss2d(g=7, H=0.7)
vossModel_0.7 = rast(vossModel_0.7$z)
threshold30 = quantile(as.matrix(vossModel_0.7), prob=0.3)
vossModel_0.7 = ifel(vossModel_0.7 > threshold30, 0, 1)
plot(vossModel_0.7)

# Lanscape with H value of 0.8
vossModel_0.8 = voss2d(g=7, H=0.8)
vossModel_0.8 = rast(vossModel_0.8$z)
threshold30 = quantile(as.matrix(vossModel_0.8), prob=0.3)
vossModel_0.8 = ifel(vossModel_0.8 > threshold30, 0, 1)
plot(vossModel_0.8)

# Lanscape with H value of 0.9
vossModel_0.9 = voss2d(g=7, H=0.9)
vossModel_0.9 = rast(vossModel_0.9$z)
threshold30 = quantile(as.matrix(vossModel_0.9), prob=0.3)
vossModel_0.9 = ifel(vossModel_0.9 > threshold30, 0, 1)
plot(vossModel_0.9)
```

Answer
As the value of H increases from 0.1 to 0.9, the simulated landscapes show a trend toward larger, more connected patches of forest cover. With a H-value of 0.1, the landscape is characterized by small, irregular patches with limited connectivity. Patches start to consolidate, and larger connected areas appear, with a 0.2 H-value. There is continued consolidation and connectivity of patches and larger forested areas dominate the landscape as the H-value increases from 0.3 to 0.6. The landscape becomes more homogeneous with even larger, connected forested areas. The patches become more regular and less fragmented with an H-value of 0.9.


**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. THESE MUST BE DIFFERENT THAN THOSE METRICS YOU USED IN CHALLENGE 2. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is H-value and y-axis is calculated metric). Briefly describe why you chose these 3 metrics and how they change with increasing cover.**

```{r}
# Place your R code here.

#Metrics; Landscape Shape Index, Mean of Patch Area and Aggregation Index

# Landscape Shape Index
LSI_0.1 = lsm_c_lsi(vossModel_0.1) %>% 
  filter(class==1)
LSI_0.1 <- LSI_0.1$value

LSI_0.2 = lsm_c_lsi(vossModel_0.2) %>% 
  filter(class==1)
LSI_0.2 <- LSI_0.2$value

LSI_0.3 = lsm_c_lsi(vossModel_0.3) %>% 
  filter(class==1)
LSI_0.3 <- LSI_0.3$value

LSI_0.4 = lsm_c_lsi(vossModel_0.4) %>% 
  filter(class==1)
LSI_0.4 <- LSI_0.4$value

LSI_0.5 = lsm_c_lsi(vossModel_0.5) %>% 
  filter(class==1)
LSI_0.5 <- LSI_0.5$value

LSI_0.6 = lsm_c_lsi(vossModel_0.6) %>% 
  filter(class==1)
LSI_0.6 <- LSI_0.6$value

LSI_0.7 = lsm_c_lsi(vossModel_0.7) %>% 
  filter(class==1)
LSI_0.7 <- LSI_0.7$value

LSI_0.8 = lsm_c_lsi(vossModel_0.8) %>% 
  filter(class==1)
LSI_0.8 <- LSI_0.8$value

LSI_0.9 = lsm_c_lsi(vossModel_0.9) %>% 
  filter(class==1)
LSI_0.9 <- LSI_0.9$value

#Plotting
Landscape_Shape_Index <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  LSI_Values = c(LSI_0.1, LSI_0.2, LSI_0.3,
                 LSI_0.4, LSI_0.5, LSI_0.6,
                 LSI_0.7, LSI_0.8, LSI_0.9)
)
 
ggplot(Landscape_Shape_Index, aes(x = Threshold, y = LSI_Values)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Landscape Shape Index vs H Value",
       x = "H-value",
       y = "Landscape Shape Index")


# Mean of Patch Area
MPA_0.1 = lsm_c_area_mn(vossModel_0.1, directions=8) %>% 
  filter(class==1)
MPA_0.1 <- MPA_0.1$value

MPA_0.2 = lsm_c_area_mn(vossModel_0.2, directions=8) %>% 
  filter(class==1)
MPA_0.2 <- MPA_0.2$value

MPA_0.3 = lsm_c_area_mn(vossModel_0.3, directions=8) %>% 
  filter(class==1)
MPA_0.3 <- MPA_0.3$value

MPA_0.4 = lsm_c_area_mn(vossModel_0.4, directions=8) %>% 
  filter(class==1)
MPA_0.4 <- MPA_0.4$value

MPA_0.5 = lsm_c_area_mn(vossModel_0.5, directions=8) %>% 
  filter(class==1)
MPA_0.5 <- MPA_0.5$value

MPA_0.6 = lsm_c_area_mn(vossModel_0.6, directions=8) %>% 
  filter(class==1)
MPA_0.6 <- MPA_0.6$value

MPA_0.7 = lsm_c_area_mn(vossModel_0.7, directions=8) %>% 
  filter(class==1)
MPA_0.7 <- MPA_0.7$value

MPA_0.8 = lsm_c_area_mn(vossModel_0.8, directions=8) %>% 
  filter(class==1)
MPA_0.8 <- MPA_0.8$value

MPA_0.9 = lsm_c_area_mn(vossModel_0.9, directions=8) %>% 
  filter(class==1)
MPA_0.9 <- MPA_0.9$value

#Plotting
Mean_Patch_Area <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  MPA_Values = c(MPA_0.1, MPA_0.2, MPA_0.3,
                 MPA_0.4, MPA_0.5, MPA_0.6,
                 MPA_0.7, MPA_0.8, MPA_0.9)
)
 
ggplot(Mean_Patch_Area, aes(x = Threshold, y = MPA_Values)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Mean Patch Area vs H Value",
       x = "H-value",
       y = "Mean Patch Area")


# Aggregation Index
AI_0.1 = lsm_l_ai(vossModel_0.1, directions=8)
AI_0.1 <- AI_0.1$value

AI_0.2 = lsm_l_ai(vossModel_0.2, directions=8)
AI_0.2 <- AI_0.2$value

AI_0.3 = lsm_l_ai(vossModel_0.3, directions=8)
AI_0.3 <- AI_0.3$value

AI_0.4 = lsm_l_ai(vossModel_0.4, directions=8)
AI_0.4 <- AI_0.4$value

AI_0.5 = lsm_l_ai(vossModel_0.5, directions=8)
AI_0.5 <- AI_0.5$value

AI_0.6 = lsm_l_ai(vossModel_0.6, directions=8)
AI_0.6 <- AI_0.6$value

AI_0.7 = lsm_l_ai(vossModel_0.7, directions=8)
AI_0.7 <- AI_0.7$value

AI_0.8 = lsm_l_ai(vossModel_0.8, directions=8)
AI_0.8 <- AI_0.8$value

AI_0.9 = lsm_l_ai(vossModel_0.9, directions=8)
AI_0.9 <- AI_0.9$value

#Plotting
Aggregation_Index <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  AI_Values = c(AI_0.1, AI_0.2, AI_0.3,
                 AI_0.4, AI_0.5, AI_0.6,
                 AI_0.7, AI_0.8, AI_0.9)
)
 
ggplot(Aggregation_Index, aes(x = Threshold, y = AI_Values)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Aggregation Index vs H Value",
       x = "H-value",
       y = "Aggregation Index")

```

Answer

Metrics used to capture changes; Landscape Shape Index, Mean of Patch Area and Aggregation Index

Why These Metrics and how they change 

Landscape Shape Index (LSI); LSI is an 'Aggregation metric', which quantifies the complexity of landscape shape. LSI = 1 when only one squared patch is present or all patches are maximally aggregated. Increases, without limit, as the length of the actual edges increases and the patches become less compact. Thus, higher values indicating more irregular and complex shapes. As H_value increases, LSI values reduce, suggesting that the landscapes become more regular and simple in their boundaries.

Mean of Patch Area (AREA_MN); AREA_MN is an 'Area and Edge metric' which summarises each class as the mean of all patch areas belonging to the class. This metric represents the average size of patches within the forest class, providing insights into the distribution of patch sizes. As H-value increases, AREA_MN increases, indicating larger and more connected forest patches. Higher H values may lead to the consolidation of forested areas into larger, more cohesive patches.

Aggregation Index (AI); AI is an 'Aggregation metric' which equals the number of like adjacencies divided by the theoretical maximum possible number of like adjacencies for that class summed over each class for the entire landscape.AI measures the degree of spatial aggregation of different land cover types, providing insights into the spatial arrangement of patches. As H increases, AI increases, suggesting more spatially aggregated and connected forest patches. Just like in AREA_MN,higher H values may lead to landscapes with larger, more cohesive forested areas.

$\color{red}{\text{Excellent work, Sinka! +4}}$
